---
title: "tfbpmodeling Paper Figures"
output: html_document
---

## Setup

Set `data_pull_date` to the date of the yeast database data version to use,
and `tfbpmodeling_version` to the version of the tfbpmodeling software that
produced the results. If `data/<data_pull_date>/` or
`data/<data_pull_date>/<tfbpmodeling_version>/` do not exist, run
`Rscript R/prepare_input_data.R` and `Rscript R/prepare_results_data_v1.R`
(or `prepare_results_data.R` for legacy results) first.

```{r setup, message=FALSE}
library(here)
library(tidyverse)
library(corrr)
library(cowplot)
library(patchwork)

data_pull_date       <- "20250805"
variant              <- "residuals"
tfbpmodeling_version <- "1.1.0"

source(here("R/prepare_data.R"))
source(here("R/fit_ols_model.R"))

plots_dir <- here("plots", data_pull_date, tfbpmodeling_version)
dir.create(plots_dir, recursive = TRUE, showWarnings = FALSE)
```

---

## Fig 1A–B: TF binding-set size distribution (FDR vs DTO)

```{r fig1ab}
source(here("R/fig1ab.R"))

tf_target_counts <- input_data$predictors_list$raw_pvalue %>%
    pivot_longer(-target_symbol,
                 names_to  = "regulator_symbol",
                 values_to = "mlog10pvalue") %>%
    group_by(regulator_symbol) %>%
    mutate(bh_fdr = p.adjust(10^(-mlog10pvalue), method = "BH")) %>%
    filter(bh_fdr <= 0.05) %>%
    ungroup() %>%
    count(regulator_symbol, name = "n") %>%
    mutate(source = "bh") %>%
    bind_rows(
        mcisaac_kem_predictor_rr %>%
            filter(expression_source == "mcisaac_oe") %>%
            select(regulator_symbol, binding_set_size) %>%
            rename(n = binding_set_size) %>%
            mutate(source = "dto")
    ) %>%
    group_by(regulator_symbol) %>%
    filter(n() == 2) %>%
    ungroup() %>%
    mutate(source = factor(source,
                           levels = c("bh", "dto"),
                           labels = c("FDR <= 0.05", "DTO")))

save_fig1ab(
    path             = file.path(plots_dir, "fig1ab.svg"),
    tf_target_counts = tf_target_counts
)
make_fig1ab(tf_target_counts)
```

---

## Fig 1B1: Replicate vs TF-TF predictor correlation

```{r fig1b1, cache=TRUE}
source(here("R/fig1b1.R"))

pval_epsilon <- 1e-300

# ---- predictor correlations ------------------------------------------------
corr_mat_and_vis <- function(df, row_order = NULL, method = "pearson") {
    data_matrix <- df %>% select(-target_symbol) %>% as.matrix()
    cor_matrix  <- cor(data_matrix, method = method)
    if (!is.null(row_order)) {
        cor_matrix   <- cor_matrix[row_order, row_order]
        cluster_rows <- FALSE; cluster_cols <- FALSE
    } else {
        cluster_rows <- TRUE; cluster_cols <- TRUE
    }
    pheatmap_res <- pheatmap::pheatmap(
        cor_matrix, clustering_method = "ward.D2",
        cluster_rows = cluster_rows, cluster_cols = cluster_cols,
        silent = TRUE
    )
    list(mat = cor_matrix, plot = pheatmap_res)
}

get_long_corr <- function(corr_mat, data_desc) {
    corr_mat[lower.tri(corr_mat)] <- NA
    corr_mat %>%
        as_tibble(rownames = "var1") %>%
        pivot_longer(-var1, names_to = "var2", values_to = "corr") %>%
        filter(!is.na(corr)) %>%
        mutate(data_transform = data_desc)
}

init_lrb  <- corr_mat_and_vis(input_data$predictors_list$log_rank_normalized)
lrb_order <- rownames(init_lrb$mat)[init_lrb$plot$tree_row$order]
cor_results <- list(
    enrichment = corr_mat_and_vis(input_data$predictors_list$raw_enrichment, row_order = lrb_order),
    pval       = corr_mat_and_vis(input_data$predictors_list$raw_pvalue,     row_order = lrb_order),
    lrb        = corr_mat_and_vis(input_data$predictors_list$log_rank_normalized, row_order = lrb_order)
)
long_corr_pearson_res <- map(names(cor_results),
                              ~ get_long_corr(cor_results[[.]]$mat, .)) %>%
    bind_rows()

# ---- replicate correlations ------------------------------------------------
spread_tied_ranks <- function(adjusted_rank) {
    N          <- length(adjusted_rank)
    ranked_min <- rank(adjusted_rank, ties.method = "min")
    ranked_avg <- rank(adjusted_rank, ties.method = "average")
    spread_rank <- numeric(N)
    for (val in sort(unique(adjusted_rank))) {
        tie_idx <- which(adjusted_rank == val)
        n_ties  <- length(tie_idx)
        if (n_ties == 1) {
            spread_rank[tie_idx] <- ranked_min[tie_idx[1]]
        } else {
            spread_rank[tie_idx] <- ranked_min[tie_idx[1]] +
                (ranked_avg[tie_idx[1]] / N) * n_ties
        }
    }
    spread_rank
}

cc_ranked <- cc_usable_data %>%
    filter(target_symbol %in% unique(input_data$predictors_list$log_rank$target_symbol)) %>%
    select(regulator_symbol, record_id, target_symbol,
           callingcards_enrichment, poisson_pval) %>%
    group_by(record_id) %>%
    mutate(primary_rank = rank(poisson_pval, ties.method = "min")) %>%
    group_by(record_id, primary_rank) %>%
    mutate(adjusted_rank = if (n() == 1) primary_rank else
        primary_rank + rank(-callingcards_enrichment, ties.method = "average") / 1e6) %>%
    ungroup() %>%
    group_by(record_id) %>%
    mutate(
        spread_rank = spread_tied_ranks(adjusted_rank),
        lrb         = -log10(spread_rank) + log10(max(spread_rank, na.rm = TRUE)),
        mlog10pval  = -log10(poisson_pval + pval_epsilon)
    ) %>%
    ungroup()

compute_replicate_correlations <- function(value_col, df) {
    df %>%
        select(regulator_symbol, record_id, target_symbol, !!sym(value_col)) %>%
        group_by(regulator_symbol) %>%
        filter(n_distinct(record_id) > 1) %>%
        ungroup() %>%
        group_by(regulator_symbol) %>%
        nest() %>%
        mutate(
            wide  = map(data, ~ pivot_wider(.x, names_from = record_id,
                                            values_from = !!sym(value_col))),
            corrs = map(wide, ~ {
                dat     <- select(.x, -target_symbol)
                pearson  <- correlate(dat, method = "pearson",
                                      use = "pairwise.complete.obs") %>%
                    shave() %>% corrr::stretch(na.rm = TRUE) %>% rename(pearson = r)
                spearman <- correlate(dat, method = "spearman",
                                      use = "pairwise.complete.obs") %>%
                    shave() %>% corrr::stretch(na.rm = TRUE) %>% rename(spearman = r)
                full_join(pearson, spearman, by = c("x", "y"))
            })
        ) %>%
        select(regulator_symbol, corrs) %>%
        unnest(corrs) %>%
        rename(record_id_1 = x, record_id_2 = y) %>%
        pivot_longer(-c(regulator_symbol, record_id_1, record_id_2),
                     names_to = "method", values_to = "corr") %>%
        mutate(data_transform = value_col)
}

replicate_corr_comp_df <- map(
    c("callingcards_enrichment", "mlog10pval", "lrb"),
    compute_replicate_correlations, cc_ranked
) %>%
    bind_rows() %>%
    ungroup() %>%
    mutate(data_transform = factor(
        data_transform,
        levels = c("callingcards_enrichment", "mlog10pval", "lrb"),
        labels = c("Abs(enrichment)", "-log10(pvalue)", "LRB")
    ))

save_fig1b1(
    path                   = file.path(plots_dir, "fig1b1.svg"),
    long_corr_pearson_res  = long_corr_pearson_res,
    replicate_corr_comp_df = replicate_corr_comp_df
)
make_fig1b1(long_corr_pearson_res, replicate_corr_comp_df)
```

---

## Fig 1C–D and 1E–H: R² across transforms, and example scatter panels

`uni_modeling_results` is shared by both fig1cd and fig1efgh.

```{r uni_modeling_results, message=FALSE, cache=TRUE}
model_regulators <- intersect(
    colnames(input_data$predictors_list$log_rank_normalized),
    colnames(input_data$responses_list$log_rank_normalized)
)
model_regulators <- model_regulators[model_regulators != "target_symbol"]

model_configs <- list(
    abs_effect_cc_effect = list(
        response  = input_data$responses_list$raw,
        predictor = input_data$predictors_list$raw_enrichment
    ),
    abs_effect_cc_pval = list(
        response  = input_data$responses_list$raw,
        predictor = input_data$predictors_list$raw_pvalue
    ),
    rank_rank = list(
        response  = input_data$responses_list$rank,
        predictor = input_data$predictors_list$rank
    ),
    logrank_logrank = list(
        response  = input_data$responses_list$log_rank_normalized,
        predictor = input_data$predictors_list$log_rank_normalized
    )
)

uni_modeling_results <- map(list(degree1 = FALSE, degree3 = TRUE), function(cubic_flag) {
    map(model_configs, function(cfg) {
        map(
            .x = set_names(model_regulators),
            .f = fit_ols_model,
            responses   = cfg$response,
            predictors  = cfg$predictor,
            cubic_model = cubic_flag
        )
    })
})
```

```{r fig1cd}
source(here("R/fig1cd.R"))

r2_summary_df <- imap_dfr(uni_modeling_results, function(degree_list, model_name) {
    imap_dfr(degree_list, function(regulator_list, data_variant_name) {
        imap_dfr(regulator_list, function(result, regulator_name) {
            fstat <- summary(result$fit)$fstatistic
            tibble(
                model        = model_name,
                data_variant = data_variant_name,
                regulator    = regulator_name,
                r2           = summary(result$fit)$r.squared,
                fstat_pval   = pf(fstat[1], fstat[2], fstat[3], lower.tail = FALSE)
            )
        })
    })
})

save_fig1cd(
    path          = file.path(plots_dir, "fig1cd.svg"),
    r2_summary_df = r2_summary_df
)
make_fig1cd(r2_summary_df)
```

```{r fig1efgh, message=FALSE}
source(here("R/fig1efgh.R"))

save_fig1efgh(
    path                 = file.path(plots_dir, "fig1efgh.svg"),
    uni_modeling_results = uni_modeling_results
)
make_fig1efgh(uni_modeling_results)
```

---

## Fig 2B–C: Bootstrap coefficient distributions (Stage 1 vs Stage 2)

```{r fig2bc}
source(here("R/fig2bc.R"))

save_fig2bc(
    path                = file.path(plots_dir, "fig2bc.svg"),
    all_data_coefs_list = bootstrap_coefs_all_data,
    topn_coefs_list     = bootstrap_coefs_topn,
    ci_df_all_data      = ci_df_all_data,
    ci_df_topn          = ci_df_topn,
    regulator           = "SKO1"
)
make_fig2bc(
    all_data_coefs_list = bootstrap_coefs_all_data,
    topn_coefs_list     = bootstrap_coefs_topn,
    ci_df_all_data      = ci_df_all_data,
    ci_df_topn          = ci_df_topn,
    regulator           = "SKO1"
)
```

---

## Fig 2D: Interactor counts per stage

```{r fig2d}
source(here("R/fig2d.R"))

save_fig2d(
    path          = file.path(plots_dir, "fig2d.svg"),
    stage_comp_df = stage_comp_df,
    stage3_col = "stage3_bootstrap_n_interactors",
    include_0_interactors = FALSE,
)

make_fig2d(stage_comp_df,
           stage3_col = "stage3_bootstrap_n_interactors",
           include_0_interactors = FALSE)
```

---

## Fig 2E: CV variance explained across stages

```{r fig2e_prep, message=FALSE, cache=TRUE}
source(here("R/fig2e.R"))

stratification_classification <- function(binding_vector,
                                           bins = c(0, 8, 64, 512, Inf)) {
    if (length(bins) < 2) {
        warning("bins < 2; returning all ones.")
        return(rep(1L, length(binding_vector)))
    }
    labels       <- seq_len(length(bins) - 1)
    binding_rank <- rank(-binding_vector, ties.method = "min")
    binding_bin  <- cut(binding_rank, breaks = bins, labels = labels, right = TRUE)
    as.integer(binding_bin)
}

significant_all_data_98_df <- ci_df_all_data %>%
    group_by(regulator_symbol = regulator) %>%
    summarise(interactors = list(interactor), .groups = "drop") %>%
    mutate(n_interactors = map_int(interactors, length))

significant_topn_90_df <- ci_df_topn %>%
    group_by(regulator_symbol = regulator) %>%
    summarise(interactors = list(interactor), .groups = "drop") %>%
    mutate(n_interactors = map_int(interactors, length))

original_response_tss <- input_data$responses_list$log_rank_normalized %>%
    pivot_longer(-target_symbol, names_to = "model", values_to = "lrr_norm") %>%
    filter(target_symbol != model) %>%
    group_by(model) %>%
    reframe(tss = var(lrr_norm) * (n() - 1))

message("Running cubic CV models...")
cubic_model_cv_r2_df <- tibble(
    model    = significant_all_data_98_df$regulator_symbol,
    var_expl = unlist(map(model, cubic_model_cv_r2)) * 100,
    stage    = "cubic_model"
)

message("Running Stage 1 (all-data) CV models...")
rss_tss_res_all_data <- map2(
    significant_all_data_98_df$regulator_symbol,
    significant_all_data_98_df$interactors,
    r2_lasso_all_data_models, cv = TRUE
) %>%
    bind_rows() %>%
    group_by(model) %>%
    reframe(sum_test_rss = sum(test_rss)) %>%
    left_join(original_response_tss, by = "model") %>%
    mutate(var_expl = (1 - sum_test_rss / tss) * 100, stage = "stage1")

message("Running Stage 2 (top-N) CV models...")
rss_tss_res_topn <- map2(
    significant_topn_90_df$regulator_symbol,
    significant_topn_90_df$interactors,
    r2_lasso_all_data_models, cv = TRUE
) %>%
    bind_rows() %>%
    group_by(model) %>%
    reframe(sum_test_rss = sum(test_rss)) %>%
    left_join(original_response_tss, by = "model") %>%
    mutate(var_expl = (1 - sum_test_rss / tss) * 100, stage = "stage2")

message("Running Stage 3 (final lasso) CV models...")
stage3_input_df <- stage3_consistent_df %>%
    group_by(model) %>%
    reframe(interactors = list(interactor))

rss_tss_res_stage3 <- map2(
    stage3_input_df$model,
    stage3_input_df$interactors,
    r2_lasso_all_data_models, cv = TRUE
) %>%
    bind_rows() %>%
    group_by(model) %>%
    reframe(sum_test_rss = sum(test_rss)) %>%
    left_join(original_response_tss, by = "model") %>%
    mutate(var_expl = (1 - sum_test_rss / tss) * 100, stage = "stage3")

stages_comp_r2_df <- cubic_model_cv_r2_df %>%
    select(model, var_expl) %>%
    rename(cubic_model = var_expl) %>%
    left_join(rss_tss_res_all_data %>% select(model, var_expl) %>%
                  rename(stage1 = var_expl), by = "model") %>%
    left_join(rss_tss_res_topn %>% select(model, var_expl) %>%
                  rename(stage2 = var_expl), by = "model") %>%
    left_join(rss_tss_res_stage3 %>% select(model, var_expl) %>%
                  rename(stage3 = var_expl), by = "model") %>%
    mutate(
        stage2 = ifelse(is.na(stage2), stage1, stage2),
        stage3 = ifelse(is.na(stage3), stage2, stage3)
    ) %>%
    pivot_longer(-model, names_to = "stage", values_to = "var_expl") %>%
    mutate(stage = factor(
        stage,
        levels = c("cubic_model", "stage1", "stage2", "stage3"),
        labels = c("Cubic Model", "Stage 1", "Stage 2", "Stage 3")
    ))
```

```{r fig2e}
save_fig2e(
    path             = file.path(plots_dir, "fig2e.svg"),
    stages_comp_r2_df = stages_comp_r2_df
)
make_fig2e(stages_comp_r2_df)
```

---

## Fig S2A: PDR3 × GAL4 cubic + interaction scatter

```{r figS2A}
source(here("R/figS2A.R"))

save_figS2A(
    path       = file.path(plots_dir, "figS2A.svg"),
    input_data = input_data,
    pTF        = "PDR3",
    mTF        = "GAL4"
)
make_figS2A(input_data, pTF = "PDR3", mTF = "GAL4")
```

---

## Fig S1A–C: TCC vs ChIP-chip binding comparison

`figS1A` uses `matplotlib_venn` via reticulate. The virtualenv must be
activated before any reticulate call fires, so it is set here.

```{r figS1-setup}
library(reticulate)
use_virtualenv(here(".venv"), required = TRUE)

source(here("R/figS1A.R"))
source(here("R/figS1B.R"))
source(here("R/figS1C.R"))

s1_ints <- .compute_s1_intersects(predictors_meta, response_meta, harb_meta, rr_meta)
```

### Fig S1A: Venn diagram (TCC vs ChIP-chip significant TFs)

```{r figS1A}
figS1A_path <- file.path(plots_dir, "figS1A.svg")
save_figS1A(figS1A_path, s1_ints$harbison, s1_ints$cc)
knitr::include_graphics(figS1A_path)
```

### Fig S1B: DTO p-value boxplot (TCC vs ChIP-chip)

```{r figS1B}
save_figS1B(
    path                  = file.path(plots_dir, "figS1B.svg"),
    harbison_rr_intersect = s1_ints$harbison,
    cc_rr_intersect       = s1_ints$cc
)
make_figS1B(s1_ints$harbison, s1_ints$cc)
```

### Fig S1C: rank_25 response rate boxplot (TCC vs ChIP-chip)

```{r figS1C}
save_figS1C(
    path                  = file.path(plots_dir, "figS1C.svg"),
    harbison_rr_intersect = s1_ints$harbison,
    cc_rr_intersect       = s1_ints$cc
)
make_figS1C(s1_ints$harbison, s1_ints$cc)
```
